---
title: "Analysis of Rental Stress Across Australian Regions and Dwelling Types in 2026"
subtitle: "ETC5513 -- Assignment3"
author: "Tin Nguyen, Waranya Ratirotjanakul, Amisha Vatsa, Inayat Gill"
date: today
format:
  revealjs:
    slide-number: true
    show-slide-number: all
    controls: true
    theme: simple
    transition: fade
    width: 1280
    height: 720
    embed-resources: true
    code-line-numbers: false
    highlight-style: github
---

## Problem introduction {.smaller}
::: smaller
**Why does rental affordability matter?**

- Australia's rental market has faced sustained pressure from population growth, low vacancy rates, and rising construction costs
- Weekly rents have moved beyond the reach of many low- to middle-income households across metropolitan and regional areas
- Understanding price variation across states and dwelling types is critical for housing policy and renters

**Research Question**

- *Where and in what types of dwellings is rental stress most acute in Australia in 2026?*
:::

## Dataset description {.smaller}

::: smaller
**Australian Rental Market Data 2026**

- Source: Kaggle
- 6,766 rental listings across Australia
- Covers all states and territories

**Information Available**

- Weekly rental price
- Geographic location
- Dwelling type
- Property characteristics (bedrooms, bathrooms, parking spaces)

**Purpose**

- Examine rental stress across Australian regions and dwelling types

```{r setup, include=FALSE}
# load all required libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
#| include: false

# load dataset
rent <- read_csv("data/australian_rental_market_2026.csv")

# clean the dataset
rent_clean <- rent |>

    # remove duplicates observations
    distinct() |>

    # retain variables relevant to analysis
    select(
        title,
        price_display,
        state,
        locality,
        propertyType,
        suburb,
        bedrooms,
        bathrooms,
        parking_spaces
    )
```
:::

## Methodology {.smaller}

::: smaller
**Data Cleaning**

- Removed 1 duplicate observation
- Excluded variables not relevant to the study
- Retained rental price, location, dwelling type, and property characteristics

**Rental Stress Definition**

- Standard housing affordability benchmark
- Rental stress = weekly rent > 30% of state median weekly household income

**Analysis**

- Compare rental stress across Australian states
- Compare rental stress across dwelling types
:::

## Exploratory Data Analysis {.smaller}

::: smaller
**Key observations**

- New South Wales contains the largest number of rental listings.
- Victoria and Western Australia also contribute substantial number of listings.
- Rental listings are represented across all Australian states and territories.

```{r}
#| echo: false
#| warning: false
#| message: false

rent_clean |>
    count(state) |>
    ggplot(
        aes(
            x = reorder(state, n),
            y = n
            )
     ) +
        geom_col(fill = "#0F6E56") +
        labs(
            title = "Distribution of Rental Properties Across Australian States",
            x = "State",
            y = "Number of Properties"
            ) +
        theme_minimal(base_size = 11)
```
:::

## Results

## Conclusions and/or recommendations

