---
title: "Analysis of Rental Stress Across Australian Regions and Dwelling Types in 2026"
subtitle: "ETC5513 -- Assignment3"
author: "Tin Nguyen, Waranya Ratirotjanakul, Amisha Vatsa, Inayat Gill"
date: today
format:
  revealjs:
    slide-number: true
    show-slide-number: all
    controls: true
    theme: simple
    transition: fade
    width: 1280
    height: 720
    embed-resources: true
    code-line-numbers: false
    highlight-style: github
---

## Problem introduction

## Dataset description {.smaller}

::: smaller
**Australian Rental Market Data 2026**

- Source: Kaggle
- 6,766 rental listings across Australia
- Covers all states and territories

**Information Available**

- Weekly rental price
- Geographic location
- Dwelling type
- Property characteristics (bedrooms, bathrooms, parking spaces)

**Purpose**

- Examine rental stress across Australian regions and dwelling types

```{r setup, include=FALSE}
# load all required libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
#| include: false

# load dataset
rent <- read_csv("data/australian_rental_market_2026.csv")

# clean the dataset
rent_clean <- rent |>

    # remove duplicates observations
    distinct() |>

    # retain variables relevant to analysis
    select(
        title,
        price_display,
        state,
        locality,
        propertyType,
        suburb,
        bedrooms,
        bathrooms,
        parking_spaces
    )
```
:::

## Methodology {.smaller}

::: smaller
**Data Cleaning**

- Removed 1 duplicate observation
- Excluded variables not relevant to the study
- Retained rental price, location, dwelling type, and property characteristics

**Rental Stress Definition**

- Standard housing affordability benchmark
- Rental stress = weekly rent > 30% of state median weekly household income

**Analysis**

- Compare rental stress across Australian states
- Compare rental stress across dwelling types
:::

## Exploratory Data Analysis {.smaller}

::: smaller
**Key observations**

- New South Wales contains the largest number of rental listings.
- Victoria and Western Australia also contribute substantial number of listings.
- Rental listings are represented across all Australian states and territories.

```{r}
#| echo: false
#| warning: false
#| message: false

rent_clean |>
    count(state) |>
    ggplot(
        aes(
            x = reorder(state, n),
            y = n
            )
     ) +
        geom_col(fill = "#0F6E56") +
        labs(
            title = "Distribution of Rental Properties Across Australian States",
            x = "State",
            y = "Number of Properties"
            ) +
        theme_minimal(base_size = 11)
```
:::

## Results

## Discussion  {.smaller style="font-size: 0.75em;"}

::: smaller
**Rental stress is widespread — but geography matters**

- Nearly **60% of all listings** exceed the 30% affordability threshold nationally
- Queensland (75.1%) has the highest stress rate, yet NSW has the largest absolute number of stressed listings (1,678)
- This divergence shows stress *rates* and *volumes* tell different stories

**Income — not rent — is the key driver**

- The ACT has the **highest median rent** but the **lowest stress rate (11.7%)**
- Its substantially higher household income shields most renters from stress classification
- Reinforces the need for income-adjusted affordability measures

**Dwelling type creates a clear affordability gradient**

- Acreage/semi-rural: highest stress (88.2%), highest rent ($970/wk) — least accessible
- Houses (65.2%) and apartments (63.1%) — high stress despite being the most common types
- Studios: lowest stress (11.5%), lowest rent ($430/wk) — most accessible option
:::

---

## Conclusion & Recommendations {.smaller}

::: smaller
::: columns
::: {.column width="50%"}
**Key Takeaways**

- Nearly **3 in 5 listings** are unaffordable for a median-income household
- Queensland, WA, and Tasmania face the greatest proportional pressure
- Income is a **stronger affordability lever** than rent levels alone
- Rental unaffordability is both **widespread and structurally entrenched**

:::

::: {.column width="50%"}
**Recommendations**

🏛️ Expand Rent Assistance calibrated to state-level incomes

🏗️ Increase supply of studios and one-bedroom apartments through rezoning

🏘️ Build-to-rent schemes and social housing investment for mainstream dwellings

📊 Future research to incorporate suburb-level income and vacancy rate data

:::
:::
:::

