---
title: "Analysis of Rental Stress Across Australian Regions and Dwelling Types in 2026"
subtitle: "ETC5513 -- Assignment3"
author: "Tin Nguyen, Waranya Ratirotjanakul, Amisha Vatsa, Inayat Gill"
date: today
format:
  revealjs:
    slide-number: true
    show-slide-number: all
    controls: true
    theme: simple
    transition: fade
    width: 1280
    height: 720
    embed-resources: true
    code-line-numbers: false
    highlight-style: github
---

## Problem introduction {.smaller}
::: smaller
**Why does rental affordability matter?**

- Australia's rental market has faced sustained pressure from population growth, low vacancy rates, and rising construction costs
- Weekly rents have moved beyond the reach of many low- to middle-income households across metropolitan and regional areas
- Understanding price variation across states and dwelling types is critical for housing policy and renters

**Research Question**

- *Where and in what types of dwellings is rental stress most acute in Australia in 2026?*
:::

## Dataset description {.smaller}

::: smaller
**Australian Rental Market Data 2026**

- Source: Kaggle
- 6,766 rental listings across Australia
- Covers all states and territories

**Information Available**

- Weekly rental price
- Geographic location
- Dwelling type
- Property characteristics (bedrooms, bathrooms, parking spaces)

**Purpose**

- Examine rental stress across Australian regions and dwelling types

```{r setup, include=FALSE}
# load all required libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
```

```{r}
#| include: false

# load dataset
rent <- read_csv("data/australian_rental_market_2026.csv")

# clean the dataset
rent_clean <- rent |>

    # remove duplicates observations
    distinct() |>

    # retain variables relevant to analysis
    select(
        title,
        price_display,
        state,
        locality,
        propertyType,
        suburb,
        bedrooms,
        bathrooms,
        parking_spaces
    )
```
:::

## Methodology {.smaller}

::: smaller
**Data Cleaning**

- Removed 1 duplicate observation
- Excluded variables not relevant to the study
- Retained rental price, location, dwelling type, and property characteristics

**Rental Stress Definition**

- Standard housing affordability benchmark
- Rental stress = weekly rent \> 30% of state median weekly household income

**Analysis**

- Compare rental stress across Australian states
- Compare rental stress across dwelling types
:::

## Exploratory Data Analysis {.smaller .scrollable}

::: smaller
**Key observations**

- New South Wales contains the largest number of rental listings.
- Victoria and Western Australia also contribute substantial number of listings.
- Rental listings are represented across all Australian states and territories.

```{r}
#| echo: false
#| warning: false
#| message: false

rent_clean |>
    count(state) |>
    ggplot(
        aes(
            x = reorder(state, n),
            y = n
            )
     ) +
        geom_col(fill = "#0F6E56") +
        labs(
            title = "Distribution of Rental Properties Across Australian States",
            x = "State",
            y = "Number of Properties"
            ) +
        theme_minimal(base_size = 11)
```
:::

## Results: Rental Stress Across Australian States {.smaller}

::: smaller
- Across the full dataset, **59.5% of listings** exceed the 30% affordability threshold
- **Queensland** recorded the highest stress rate (75.1%), followed by Western Australia (71.6%) and Tasmania (67.3%)
- Despite having the **highest median rent**, the ACT had the lowest stress rate (11.7%) — reflecting its substantially higher household income

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-stress-state
#| fig-cap: "Proportion of rental listings exceeding the 30% income threshold (rental stress) by Australian state, 2026."

state_income <- tibble(
    state = c("NSW", "Vic", "Qld", "WA", "SA", "ACT", "Tas", "NT"),
    weekly_income = c(2142, 2011, 1943, 2197, 1847, 2817, 1654, 2109)
)

rent_clean |>
    left_join(state_income, by = "state") |>
    mutate(
        stress_threshold = weekly_income * 0.30,
        in_stress = price_display > stress_threshold
    ) |>
    group_by(state) |>
    summarise(stress_rate = mean(in_stress, na.rm = TRUE) * 100) |>
    ggplot(aes(x = reorder(state, stress_rate), y = stress_rate, fill = stress_rate)) +
    geom_col(show.legend = FALSE) +
    scale_fill_gradient(low = "#9FE1CB", high = "#0F6E56") +
    geom_hline(yintercept = 30, linetype = "dashed", colour = "grey40", linewidth = 0.5) +
    annotate("text", x = 1, y = 32, label = "30% threshold", hjust = 0,
             size = 3, colour = "grey40") +
    coord_flip() +
    labs(
        x = "State",
        y = "Listings in rental stress (%)"
    ) +
    theme_minimal(base_size = 11)
```
:::

## Results: Rental Stress Across Dwelling Types {.smaller .scrollable}

::: smaller

- **Acreage/semi-rural** properties had the highest stress rate (88.2%) and median rent (\$970/wk)
- **Houses** (65.2%) and **apartments** (63.1%) — the most common types — also recorded high stress
- **Studios** were the most affordable: 11.5% stress rate and \$430/wk median rent
:::


```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stress-type
#| tbl-cap: "Rental stress rate and median weekly rent by dwelling type."

state_income <- tibble(
    state = c("NSW", "Vic", "Qld", "WA", "SA", "ACT", "Tas", "NT"),
    weekly_income = c(2142, 2011, 1943, 2197, 1847, 2817, 1654, 2109)
)

rent_clean |>
    left_join(state_income, by = "state") |>
    mutate(
        stress_threshold = weekly_income * 0.30,
        in_stress = price_display > stress_threshold
    ) |>
    group_by(propertyType) |>
    summarise(
        `Number of listings` = n(),
        `Median rent ($/wk)` = round(median(price_display, na.rm = TRUE), 0),
        `Stress rate (%)` = round(mean(in_stress, na.rm = TRUE) * 100, 1)
    ) |>
    arrange(desc(`Stress rate (%)`)) |>
    rename(`Property type` = propertyType) |>
    knitr::kable()
```


## Conclusion {.smaller}

::: smaller
**Key takeaways from the analysis**

- Rental stress in Australia in 2026 is **the norm, not the exception**
- Nearly **3 in 5 listings** are unaffordable for a median-income household
- Queensland, Western Australia, and Tasmania face the greatest proportional pressure
- The ACT demonstrates that **income is a more powerful affordability lever than rent levels alone**

**Dwelling type findings**

- Acreage and semi-rural properties are well beyond reach for most renters
- Even mainstream options — houses and apartments — are increasingly unaffordable
- Studios offer the most accessible entry point, but represent a small share of listings

> Rental unaffordability in Australia in 2026 is both **widespread and structurally entrenched**, varying meaningfully by geography and dwelling type
:::

## Recommendations {.smaller style="font-size: 0.75em;"}

::: smaller
**Policy Priorities for Australia's Rental Market**

🏛️ **Target high-stress states**

- Expand Commonwealth Rent Assistance calibrated to state-level incomes
- Priority states: Queensland, Western Australia, and Tasmania

🏗️ **Increase affordable supply**

- Prioritise studios and one-bedroom apartments in high-stress states
- Enable growth through rezoning and density incentives

🏘️ **Address mainstream housing**

- Invest in build-to-rent and social housing schemes
- Introduce inclusionary zoning for new house and apartment developments
:::

::: {.column width="50%"}
**Recommendations**

🏛️ Expand Rent Assistance calibrated to state-level incomes

🏗️ Increase supply of studios and one-bedroom apartments through rezoning

🏘️ Build-to-rent schemes and social housing investment for mainstream dwellings

📊 Future research to incorporate suburb-level income and vacancy rate data
:::

