---
title: "An Analysis of Rental Stress Across Australian Regions and Dwelling Types in 2026"
subtitle: "ETC5513 -- Assignment3"
author: "Tin Nguyen, Waranya Ratirotjanakul, Amisha Vatsa, Inayat Gill"
date: today
format: 
    html:
        theme: cosmo
        toc: true
        smooth-scroll: true
    typst: 
        documentclass: article
        toc: true
    
bibliography: references.bib
reference-section-title: References
---

```{r setup, include=FALSE}
# load all required libraries
library(tidyverse)
library(lubridate)
library(ggplot2)
```

# Executive summary

# Introduction 

# Methodology

## Dataset Introduction 
This study uses the Australian Rental Market Data 2026 dataset obtained from Kaggle [@kanchana2026]. The dataset provides a national snapshot of residential rental listings across Australia in 2026 and includes information on weekly rental prices, dwelling characteristics, and geographic attributes. To investigate rental price pressure across regions and dwelling types, variables related to rental price, location, dwelling category, and property characteristics were retained for analysis.

Before analysis, the dataset was reviewed for missing values, duplicate observations, and variable suitability. One duplicated observation was identified and removed to avoid repeated records. Missing values occurred in variables outside the scope of this study and therefore were not treated. Variables unrelated to the research objective, including listing text and detailed address information, were excluded to reduce unnecessary complexity. The final cleaned dataset retained variables describing rental price, geographic location, dwelling type, and property characteristics.

```{r}
#| include: false

# load dataset
rent <- read_csv("dataset/australian_rental_market_2026.csv")

# inspect dataset
glimpse(rent)

# check missing values
colSums(is.na(rent))

# check duplicates
sum(duplicated(rent))

# inspect duplicated rows
rent |> filter(
    duplicated(rent) |
    duplicated(rent, fromLast = TRUE)
)

# clean the dataset
rent_clean <- rent |>

    # remove duplicates observations
    distinct() |>

    # retain variables relevant to analysis
    select(
        title,
        price_display,
        state,
        locality,
        propertyType,
        suburb,
        bedrooms,
        bathrooms,
        parking_spaces
    )

# inspect cleaned data
glimpse(rent_clean)
```

## Dataset Description
The cleaned dataset contains `r nrow(rent_clean)` observations and `r ncol(rent_clean)` variables. The retained variables include numeric measures describing rental price and dwelling characteristics (`price_display`, `bedrooms`, `bathrooms`, and `parking_spaces`) and categorical variables describing location and dwelling category (`state`, `locality`, `suburb`, and `propertyType`). ogether, these variables support comparisons of rental prices across regions and dwelling types in Australia in 2026.

## Dataset Summary 
@tbl-property-summary summarises the composition of dwelling types and their price characteristics in the cleaned Australian rental dataset for 2026. Houses accounted for the largest share of observations (4,089 properties), followed by apartments (1,369 properties) and units (761 properties), indicating that the dataset primarily represents conventional residential dwellings. Median weekly rents varied across dwelling categories, ranging from AUD 430 for studios to AUD 970 for acreage and semi-rural properties, illustrating the diversity of prices represented in the dataset.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-property-summary
#| tbl-cap: "Summary statistics of rental prices by dwelling type."

rent_clean |>
    group_by(propertyType) |>
    summarise(
        `Number of Properties` = n(),
        `Mean Weekly Rent (AUD)` = round(mean(price_display, na.rm = TRUE), 1),
        `Median Weekly Rent (AUD)` = round(median(price_display, na.rm = TRUE), 1)
        ) |>
    arrange(desc(`Median Weekly Rent (AUD)`)) |>
    knitr::kable()
```

In addition to dwelling type characteristics, @fig-state-distribution shows the distribution of rental properties across Australia in 2026. Most observations were from New South Wales, followed by Victoria and Western Australia, indicating broad geographic coverage of the Australian rental market for subsequent analysis.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of rental properties across Australian states"

rent_clean |>
    count(state) |>
    ggplot(
        aes(
            x = reorder(state, n),
            y = n
            )
     ) +
        geom_col(fill = "#0F6E56") +
        labs(
            x = "State",
            y = "Number of Properties"
            ) +
        theme_minimal(base_size = 11)
```

# Results

Rental stress is defined here as weekly rent exceeding 30% of the median weekly household income for each state, following the standard affordability benchmark [@aihw2025]. Across the full dataset, 59.5% of listings meet this threshold, indicating that the majority of available rentals in 2026 are unaffordable for a median-income household.

@fig-stress-state shows the proportion of listings classified as rental stress by state. Queensland recorded the highest stress rate (75.1%), followed by Western Australia (71.6%) and Tasmania (67.3%). Despite having the highest median weekly rent, the Australian Capital Territory had the lowest stress rate (11.7%), reflecting its substantially higher median household income relative to other states. New South Wales, while recording a moderate stress rate of 62.2%, had the largest absolute number of stressed listings (1,678 properties) due to its volume of listings.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-stress-state
#| fig-cap: "Proportion of rental listings exceeding the 30% income threshold (rental stress) by Australian state, 2026."

state_income <- tibble(
    state = c("NSW", "Vic", "Qld", "WA", "SA", "ACT", "Tas", "NT"),
    weekly_income = c(2142, 2011, 1943, 2197, 1847, 2817, 1654, 2109)
)

rent_clean |>
    left_join(state_income, by = "state") |>
    mutate(
        stress_threshold = weekly_income * 0.30,
        in_stress = price_display > stress_threshold
    ) |>
    group_by(state) |>
    summarise(stress_rate = mean(in_stress, na.rm = TRUE) * 100) |>
    ggplot(aes(x = reorder(state, stress_rate), y = stress_rate, fill = stress_rate)) +
    geom_col(show.legend = FALSE) +
    scale_fill_gradient(low = "#9FE1CB", high = "#0F6E56") +
    geom_hline(yintercept = 30, linetype = "dashed", colour = "grey40", linewidth = 0.5) +
    annotate("text", x = 1, y = 32, label = "30% threshold", hjust = 0,
             size = 3, colour = "grey40") +
    coord_flip() +
    labs(
        x = "State",
        y = "Listings in rental stress (%)"
    ) +
    theme_minimal(base_size = 11)
```

## Rental stress by dwelling type

@tbl-stress-type summarises the rental stress rate and median weekly rent for each dwelling type. Acreage and semi-rural properties recorded the highest stress rate (88.2%) and the highest median rent (\$970/wk), making them the least affordable dwelling category. Houses (65.2%) and apartments (63.1%) followed, representing the most common dwelling types in the dataset. Studios recorded the lowest stress rate (11.5%) and the lowest median rent (\$430/wk), making them the most accessible option for renters on median incomes.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stress-type
#| tbl-cap: "Rental stress rate and median weekly rent by dwelling type."

state_income <- tibble(
    state = c("NSW", "Vic", "Qld", "WA", "SA", "ACT", "Tas", "NT"),
    weekly_income = c(2142, 2011, 1943, 2197, 1847, 2817, 1654, 2109)
)

rent_clean |>
    left_join(state_income, by = "state") |>
    mutate(
        stress_threshold = weekly_income * 0.30,
        in_stress = price_display > stress_threshold
    ) |>
    group_by(propertyType) |>
    summarise(
        `Number of listings` = n(),
        `Median rent ($/wk)` = round(median(price_display, na.rm = TRUE), 0),
        `Stress rate (%)` = round(mean(in_stress, na.rm = TRUE) * 100, 1)
    ) |>
    arrange(desc(`Stress rate (%)`)) |>
    knitr::kable()
```

# Discussion

The findings reveal a rental market under considerable strain. With nearly 60% of all listings exceeding the 30% affordability threshold, rental stress reflects a structural mismatch between housing costs and household incomes across most of Australia. Queensland's stress rate of 75.1% is particularly striking, surpassing even New South Wales despite NSW carrying the largest absolute number of unaffordable listings (1,678 properties). This divergence suggests that stress rates and raw listing volumes tell different stories — Queensland renters face a more acute affordability problem proportionally, while NSW presents the greatest overall scale of housing pressure [@aihw2025].

The ACT stands out as a clear outlier. Despite recording the highest median rent among all states, its stress rate of just 11.7% demonstrates that rent levels alone do not determine affordability — household income plays an equally decisive role. Because the 30% threshold is applied relative to each state's median weekly income, the ACT's substantially higher income base shields most renters from stress classification, even at elevated rent levels. This reinforces the importance of income-adjusted affordability measures over simple rent comparisons [@aihw2025].

At the dwelling level, the results reveal a clear affordability gradient. Acreage and semi-rural properties recorded the highest stress rate (88.2%) and the highest median rent (\$970/week), making them the least accessible category for median-income renters. Houses (65.2%) and apartments (63.1%) also recorded high stress rates, indicating that affordability pressure is concentrated in the most common dwelling types. Studios recorded the lowest stress rate (11.5%) and the lowest median rent (\$430/week), making them the most accessible option in the dataset. As this study is limited to listed rental properties, conclusions about broader supply availability cannot be drawn from these figures alone.

# Conclusion

This study examined rental affordability across Australia in 2026 using a dataset of residential listings, applying the standard 30% income-to-rent benchmark to assess rental stress at both the state and dwelling-type level [@aihw2025].

The results paint a sobering picture. Nearly three in five rental listings nationally exceed the affordability threshold, confirming that rental stress is the norm rather than the exception in the current Australian market. At the state level, Queensland, Western Australia, and Tasmania emerged as the most financially pressured regions — not necessarily because their rents are the highest, but because their household incomes are comparatively lower, making even moderate rents unaffordable for median-earning households [@abs2024]. The ACT's low stress rate (11.7%) despite its high median rent reinforces this point clearly: income is the more powerful determinant of affordability than rent price alone.

At the dwelling level, the findings reveal a clear affordability gradient. Acreage and semi-rural properties are the least accessible, with an 88.2% stress rate and a median rent of \$970 per week — placing them well beyond reach for most renters. Conventional dwellings such as houses and apartments, which make up the bulk of the market, also recorded high stress rates (65.2% and 63.1% respectively), suggesting that even mainstream housing options are increasingly unaffordable. Studios stand out as the most accessible dwelling type, with a stress rate of just 11.5% and a median rent of \$430 per week, though their small share of total listings limits their impact as a broad solution.

Taken together, these findings indicate that rental unaffordability in Australia in 2026 is both widespread and structurally entrenched, varying meaningfully by both geography and dwelling type.

# Recommendations

Based on these findings, several policy directions warrant consideration.

First, state governments in Queensland, Western Australia, and Tasmania should be prioritised for targeted rental relief measures. Given that stress in these states is driven largely by the income–rent gap rather than extreme rent levels alone, income-side interventions — such as expanded Commonwealth Rent Assistance payments calibrated to state-level median incomes — are likely to be more effective than rent caps applied uniformly at the national level [@grattan2023].

Second, increasing the supply of smaller, more affordable dwelling types — particularly studios and one-bedroom apartments — should be a planning priority in high-stress states. The data clearly show studios offer the most accessible entry point into the rental market, yet they represent a small fraction of total listings. Rezoning and density incentives in well-connected urban areas could help grow this segment of supply [@reia2024].

Third, the affordability crisis in conventional housing categories (houses and apartments) demands attention given their dominance in the dataset. Policy levers such as build-to-rent schemes, social housing investment, and inclusionary zoning requirements for new developments could help moderate stress rates in these high-volume categories over the medium term [@grattan2023].

Finally, this study highlights the limitations of applying a single national income benchmark to assess rental stress. Future research should incorporate suburb-level or postcode-level income data — available through the ABS — to produce more granular affordability assessments [@abs2024]. Integrating vacancy rates, rental growth trends, and household composition data would further strengthen the analytical basis for evidence-based housing policy in Australia.


